Partitioning growth-dependent from growth-independent influences on tropical tree mortality

### 2018-08-10 ms/manuscript.Rnw
Incorporate Mark's and Daniel's changes

\title{Unify intra- and inter-specific drivers of tropical tree mortality via additive hazards approach}

@@

% Please add here a significance statement to explain the relevance of your work
% 120 words
\significancestatement{We build a mortality model that not only integrates intra and inter-specific variation but also, simultaneously estimates the two fundamental hazards faced by plants -- Death due to low growth rates and death independent of growth -- and how a species' light demand, maximum dbh and wood density affect each. We parameterize this model using over 400,000 observed survival records collected over a 15-year period from more than 180,000 individuals across 203 species. We show that growth-dependent hazards are a significant contributor to tree mortality at Barro Colorado Island. Moreover, we show that species traits -- wood density, maximum dbh and light demand -- all affect tree mortality rates but ultimately only account for a small fraction of the total variability among species.}

@@

% 250 words
\begin{abstract}
Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified a variety of factors influencing tree mortality at both individual and species scales, these distinct drivers are yet to be integrated within a unified predictive framework. In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive hazard functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama. Specifically, we develop an empricial mortality model that not only integrates individual, species, and temporal effects, but is also the first to simultaneously estimate the independent contributions of growth-related hazards and growth-independent hazards on the overall mortality rates. We show that functions that separate mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and incorporate a species' wood density, maximum dbh and light demand outperform alternative formulations. Moreover, we show that while these three traits do affect both growth-dependent and growth-independent hazards, they account for only a small fraction of the total variability observed among species. Combined, this study provides a statistical framework for predicting individual-level mortality in highly diverse tropical forests. It also highlights how little we know about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
\end{abstract}

@@

Tree death is a key process driving population dynamics, nutrient cycling, and evolution within plant communities. While the risk of death varies among forests and from year-to-year \cite{vanMantgem:2009bi}, trees growing within the same forest also vary remarkably in realised risk of death\cite{Condit:2006}. In tropical forests, annual mortality can differ 20-60 fold among populations of coexisting species \cite{Condit:2006}. These differences reflect differences in the way trees from different species are constructed and function, and arise via a coordinated axis of life-history variation, in which individuals from species with higher potential-growth rates also experience higher mortality \cite{Wright:2010fl}. At the same time, the risk of death varies widely within species, with those individuals growing slower or sitting lower in the light spectrum at greater risk \cite{Kobe:1995tw, Wyckoff:2000un, Vieilledent:2010fv}. While these inter- and intra-specific patterns of mortality are widely recognised, they show conflicting relationships with tree growth rate (increasing with growth rate across species, decreasing with growth rate within species), making it difficult to integrate both sources of variation into a predictive model.

In this article we outline a new approach for predicting mortality among individuals within a forest, incorporating both inter and intra-specific variation in risk. Our approach builds on the notion that the total risk of death at any time is the additive effect of at least two, largely independent risks \cite{Keane:2001db}. Although plants die because of many causes, these broadly fall into two main categories \cite{Keane:2001db, McDowell:2011dr}. The first are growth-independent hazards, which cause death irrespective of the tree's current growth rate, and may include windfall, fire, and certain types of pathogen or herbivore attack. These hazards are assumed to occur at some average rate (Fig 1a, left). The second are growth-dependent hazards, which cause slower growing plants within each species to die at a faster rate, presumably because inferior carbon budgets make them unable to tolerate diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}. Many empirical studies suggest an exponential decline in risk with increased growth rate, $X_i(t)$ \cite{Kobe:1995tw, Wyckoff:2000un, Vieilledent:2010fv} (Fig 1a, centre). An individual's total risk of dying is then the sum of these two (Fig 1a, right). Biologically, the additive model implies that even fast-growing individuals experience some risk of death.

While already applied in some vegetation models \cite{Moorcroft:2001ws}, the additive hazards model has not previously been applied in empirical studies. Moreover, we note that the underlying mathematical structure of statistical methods used in most previous empirical studies imply responses similar to either the left or centre panels of Fig 1a (see SI Appendix S1). In particular, commonly used methods -- including exponential models \cite{Kobe:1995tw, Wyckoff:2000un}, logistic regression \cite{Ruger:2011cv}, generalized linear models \cite{Visser_functional_2016}, and Cox proportional-hazard models \cite{Neumann_climate_2017} -- handle covariates in a manner that allows predicted risk to decline to zero for fast growing individuals (as in Fig 1a centre; SI Appendix S1). These methods may therefore fail to capture the full range of variation observed within and among species.

@@

In forests, mortality rates are usually estimated from census data, where the size and fate of individual trees is recorded at successive time points \cite{Anderson-Teixeira-2015a} (Fig 1b). The size and status of individuals at each census -- alive vs. dead -- can then be used to estimate $\lambda_i(t)$, a continuous, latent, mortality (or `hazard`) rate for all individuals in group $i$ at time $t$. Using this approach, the probability an individual dies between times $t_1$ and $t_2$ results from the cumulative hazard they are exposed to between censuses\cite{Kleinbaum:2012tn}:
\begin{equation} \label{eq:mortality_model}
p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t\right),
\end{equation}
The observed survival outcome $S_{i, t_1\rightarrow t_2}$ (0 = alive, 1 = died) is then modelled as a binary draw. The challenge is then to estimate an appropriate form for the hazard rate, $\lambda_i(t)$.

Here we propose the following hazard function, incorporating individual-, species-, and temporal effects:
\begin{equation} \label{eq:model}
\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s}}^{\text{Growth independent}} + \overbrace{\alpha_{s} \, \textrm{e}^{-\beta_{s} \, X_{s,i}(t)}}^{\text{Growth dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
\end{equation}
Eq. \ref{eq:model} allows for additive growth-dependent and growth-independent effects, includes a negative exponential effect of individual's growth rate (our measure of carbon budget), allows the overall mortality rate to vary among censuses via the random effect $\delta_{t}$, while also allowing species-level differences in the shape of $\lambda_i(t)$ via the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$. The species-level parameters are in turn modelled as random effects, drawn from  an overall distribution.

\begin{figure}[!hb]
\centering
\includegraphics{figures/fig1.pdf}
\caption{Outline of methodology.
A) We consider three alternative hazard functions: 1) a growth-independent baseline hazard (A, left), 2) a growth-dependent hazard, where growth rate is a proxy for carbon budget (A, middle); and 3) a function that combines both growth-dependent and independent hazards (A, right). The parameters of the models are biologically interpretable: $\alpha$ defines the low growth rate (i.e. low carbon budget) effect on the overall hazard; $\beta$ reflects the sensitivity of mortality rate to changes in growth rate; and $\gamma$ is the asymptote. In the most complex model (Panel A; right), $\alpha$ and $\beta$ capture growth-dependent hazards, while $\gamma$ captures growth-independent hazards that kill a plant, irrespective of its growth rate/carbon budget (e.g. windfall, fire). For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
B) Our data consist of repeat measures of stem diameter ($D$) and status ($S$, 0 = alive or 1 = dead) for individual trees at specific census dates ($t_1$, $t_2$, $t_3$).
C) Each model's skill in predicting observed outcomes ($S$) is quantified via the log-loss function (eq. \ref{eq:logloss}). Lower logloss means higher predictive accuracy.
D) The predictive skill of alternative models was evaluated via 10-fold cross validation. The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$). Alternative models were fit 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing (1 fold; orange) and training (9 folds; green) data. Predictive capacity was assessed by averaging the log loss's obtained from the 10 test data predictions.}
\label{fig1}
\end{figure}

\section*{Case study: Tropical forest at Barro Colorado Island}

To illustrate the proposed mortality model, we assess it's ability to predict the survival status of \Sexpr{n_inds(BCI_training_full)} individual trees from containing \Sexpr{n_spp(BCI_training_full)}  species, recorded over three 5-year census periods at Barro Colorado Island (BCI), Panama. Mortality rates are known to vary widely among species, and also vary within species in response to light levels and growth rates \cite{Condit:2006, Wright:2010fl, Ruger:2011cv}. Applying the additive model, we: i) estimate growth-dependent and growth-independent effects from long-term census data, ii) estimate species-level differences in these effects, and iii) assess how well strength species-level differences is predicted from three key traits, previously shown to correlate with mortality. Here we consider three species traits thought to drive inter-specific variation in mortality: wood density, light demand, and maximum dbh \cite{king:2006va, Kraft:2010kq, Wright:2010fl, Ruger:2011cv}. Overall, these traits only show weak correlation (r$^2$ \textless 0.13) and thus provide relatively independent measures of a species ecological strategy (SI Appendix, Fig. S6).

Our analysis proceeds in a series of stages, in which we compare the predictive power of alternative models, differing in either structure or covariates included. We used 10-fold cross-validation to evaluate the skill of different models in their ability to predict outcomes in novel data (i.e. not used in model fitting), using a likelihood-based penalty for incorrect prediction (Fig. \ref{fig1}C-D; see Methods) \cite{Hooten:2015, Stone:1997tg}. n total \Sexpr{n_obs(BCI_training_full)} observations were used to fit these models. Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations. We therefore fit models across six iterative stages of model development, described below, gradually building up in complexity, comparing their predictive skill on new data. We then use the final model to assess how traits align with different effects and quantify the contributions of different effects on overall model performance.

@@

\subsection*{Stage 1: Mortality over time}
We first assess whether average mortality rates varied among the three censuses, and thus, whether a census effect is required in Eq \ref{eq:model}. We compared a model with a constant mortality rate, $\gamma$, to a model where the constant is scaled by a census effect, $\gamma\delta_{t}$. Examining the three 5-year census intervals, we found average mortality rates declining over time: with the highest proportion of death occurring from 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_training_full, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_training_full, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_training_full, 3)}\%). Consequently, we observed a small increase in predictive skill when a census effect was included (Fig. \ref{fig2}).

\subsection*{Stage 2: Hazard functions}
Second, we assessed the predictive performance of three alternative hazard functions (Fig. \ref{fig1}A). The simplest form (Fig. \ref{fig1}A, left) assumes a constant hazard. The second form (Fig. \ref{fig1}A, middle) assumes the risk of dying declines towards an asymptote of zero as growth, and thus carbon budget, increases. The third "additive" model (Fig. \ref{fig1}BA right) is the sum of the previous two. In doing this, we also investigated which growth measure (stem diameter, stem area increment) more skilfully predicts growth-dependent mortality. Comparing the three hazard functions, the additive function significantly outperformed both the baseline and growth-dependent only hazard functions (Fig. \ref{fig2}). Moreover, we found that predictive skill was higher when using stem diameter growth relative to stem-area growth (Fig. \ref{fig2}).

\subsection*{Stages 3-5: Species traits}

In the following 3 stages stages, we compare the most predictive hazard function and growth rate measure from stage 2 (additive model using dbh growth rate) and ask how addition of 1, 2 and 3 species-level traits improves model performance. Specifically, we introduce effects of the traits maximum dbh ($\psi$), light demand ($\upsilon$) and wood density ($\rho$) on the parameters $\alpha$, $\beta$ and $\gamma$ in Eq. \ref{eq:model}, at first one at a time (stage 3), then in 2-trait (stage 4) and 3-trait (stage 5) combinations. In general we found that including trait effects on these parameters decreased logarithmic loss, and thus, increased predictive accuracy (Fig.~\ref{fig2}). Moreover, we found that incorporating the effect of all three traits on these parameters resulted in higher predictive capacity relative to models that only incorporated one or two of the traits.

\subsection*{Stages 6: Species random effects}

Lastly, we fit a model that included a species random effect on each parameter. This allowed us to 1) examine the maximum predictive capacity when accounting for species differences; and 2) provides a relative indication on how much of the total species variation is captured by the three traits. We found that the increase in predictive accuracy afforded by traits (stage 5) was small relative to including species random effects on each parameter, which account for unexplained species differences, signifying that substantial inter-specific variation in mortality rates remains unexplained (Fig. \ref{fig2} Stage 6).

\subsection*{Parameter importance and trait effects}

In order to examine how species differed in hazards functions and to understand the effects of species traits, we used the entire dataset to fit final model containing all three species traits and a species random effect (which captures inter-specific variation that unexplained by traits) on each of $\alpha$, $\beta$ and $\gamma$. We found that each trait influenced tree mortality in different ways (Fig. \ref{fig3}-\ref{fig4}). Specifically, we found that light demanding species were more susceptible to low growth rates (i.e. higher $\alpha$), had higher growth-independent hazards (i.e. higher $\gamma$) and exhibited shallower declines in growth-dependent hazards with increasing growth rates (i.e. lower $\beta$). By contrast, we found that wood density had significant impacts on $\alpha$ and $\gamma$, such that species with high wood density were less susceptible to low growth rates and had lower growth-independent hazards. Species' maximum dbh only had significant effects on $\gamma$, whereby species that can achieve larger dbh exhibited lower growth-independent hazards.

@@

Using the final model, we also examine: 1) its ability distinguish between individuals that died and individuals that survived via AUROC; and 2) the amount of inter-specific variation the model explains (i.e. $r^2$) relative to the proportions of individual deaths observed per species between 1995 and 2010 (Table.~\ref{table:1}). We found that the full model had an AUCROC of \Sexpr{table1[[1,3]]}, indicating that the model had a fair ability to correctly discriminate observed deaths from survivals. Moreover, we found that the full model also explained \Sexpr{table1[[1,2]]*100}\% of the total variation observed among species.

We then examined the relative contribution species traits, random effects and growth-dependent and independent hazards (Table.~\ref{table:1}) had on overall hazard rates by individually removing their effects and examining changes in both $r^2$ and AUROC. First, we found that the removal of census random effect $\delta_t$ had little impact on either $r^2$ or AUROC. The removal of either wood density or light demand resulted in similar magnitude effects on both $r^2$ and AUROC. By contrast, the removal of maximum dbh had a slightly smaller effect on both metrics. We detected a significant drop in both $r^2$ and AUROC when species random effects (i.e. the parameters that capture variation that is otherwise unexplained by traits) were removed, further highlighting that maximum dbh, light demand and wood density explain only a small fraction of the overall inter-specific variation observed among species. As expected, the removal of both species traits and species random effects resulted in a model that was unable to capture any inter-specific variation observed among species, and thus had low discriminatory ability. Lastly, we found the removal of the growth-independent hazard term ($\gamma_s$) resulted in a lower $r^2$ relative to the removal of the growth-dependent hazard ($\alpha_s e^{-\beta_s X_{i}}$), indicating that the former hazard explained more inter-specific variation. However, when examining the removal of these hazards in terms of the model's ability to correctly distinguish survivors from those that died (AUROC), we found the opposite -- that the removal of the growth-dependent hazard resulted in substantially lower predictive ability.

@@

\section*{Discussion}

Our Bayesian framework coupled with cross validation revealed that the most predictive hazard model of tropical tree death was one which: 1) partitioned mortality into growth-dependent and growth-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) included the effects of species' maximum dbh, wood density and light demand; and 4) incorporated temporal variability. Moreover, we found that rates of tropical tree mortality varied substantially between species and that the a species' maximum dbh, wood density and light demand explained only a fraction of the overall inter-specific variation.

There are several distinct advantages of modelling tree mortality as a function of two additive, and thus independent hazards: 1) The model explicitly accounts for the independent effects of the two fundamental types of hazards faced by plants; 2) it allows for the examination of the relative contributions each hazard provides to the overall hazard rate; 3) it allows for examination of how various factors affect each hazard term, and because of this; 4) allows ecological theories of how factors such as species traits affect plant mortality to be empirically tested using commonly collected data.

The findings of this study provide empirical support for dynamic vegetation models that predict tree mortality as a function of both growth-dependent and independent hazards \cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}. We show that regardless of growth measure (our proxy for carbon budget), incorporating both hazards significantly improves model predictive skill. This is because the growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stresses (e.g. competition, parasites, herbivory). By contrast, the growth-independent hazard accounts for deaths caused by events that arise irrespective of an individual's carbon budget (e.g. windthrow, lightning strike).

Additionally, the partitioning of mortality rates into both growth-dependent and growth-independent effects allowed us to not only estimate their relative contribution, but also how species traits affect each. Like many other studies \cite{King:1994vg,king:2006va, Ruger:2011cv, Kunstler:2016km}, our analyses highlighted the importance of light competition in influencing tropical tree demographic rates. Specifically, we found that that the removal of the growth-dependent hazard term resulted in a substantial reduction in the model's ability to correctly distinguish survivors from those that died (Table~\ref{table:1}). This suggests deficiencies in carbon budgets are a major contributor to tree death on Barro Colorado Island.

Our analyses support previous research that advocated that wood density, species maximum size and light demand important explanatory variables of inter-specific variation in tropical rainforests \cite{king:2006va,Poorter:2006tq, Chave:2009iy}. Specifically, we show that light demanding species are not only more susceptible to both low growth rates and growth-independent hazards, but also exhibited shallower declines in growth-dependent hazards with increasing growth rates. By contrast, both high dbh or wood density species were found to be more tolerant of growth-independent hazards, corroborating evidence reported elsewhere \cite{Chave:2009iy, Kraft:2010kq}. However, our analyses has also supports other work \cite{vanGelder:2006exa, Kunstler:2016km} which highlights that higher wood density species are also more tolerant of low growth rates, and thus, better able to tolerate shade and competiton.

>>>>

\title{Partitioning growth-dependent from growth-independent influences on tropical tree mortality}

@@

% Please add here a significance statement to explain the relevance of your work
% 120 words
\significancestatement{We present a model that partitions rates of tropical tree mortality into growth-dependent and growth-independent hazards. This creates the opportunity to examine the relative contributions of within-species and across-species variation on tropical tree mortality rates, but also, how species traits affect each hazard. We parameterize this model using \textgreater 400,000 observed survival records collected over a 15-year period at Barro Colorado Island from more than 180,000 individuals across 203 species. We show that marginal carbon budgets are a major contributor to tree death on Barro Colorado Island. Moreover, we found that while species' light demand, maximum dbh and wood density affected tree mortality in different ways, they explained only a small fraction of the total variability observed among species.}

@@

% 250 words
\begin{abstract}
Tree death drives population dynamics, nutrient cycling, and evolution within plant communities. Mortality variation across species is thought to be influenced by different factors relative to variation within species. The unified model provided here separates mortality rates into growth-dependent and growth-independent hazards. This model creates the opportunity to simultaneously estimate the descriptors of variation both across and within species. Moreover, it provides the ability to examine how species traits affect growth-dependent and growth-independent hazards. We derive this unified mortality model  using cross-validated Bayesian methods coupled with mortality data collected over three census intervals for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama. We found that growth independent mortality tended to be higher in species with lower wood density, higher light requirements and smaller maximum dbh. Mortality due to marginal carbon budget as measured by near-zero growth rate tended to be higher in species with lower wood density and higher light demand. The total mortality variation attributable to differences among species was large relative to variation explained by these traits, emphasizing that much remains to be understood. This statistical framework strengthens our capacity to parse and understand individual-level mortality in highly diverse tropical forests, and hence to predict its consequences. 
\end{abstract}

@@

Tree death is a fundamental process affecting population dynamics, nutrient cycling, and evolution within plant communities. While the risk of death varies substantially among forests and from year to year \cite{vanMantgem:2009bi}, trees growing within the same forest also vary remarkably in realised risk of death\cite{Condit:2006}. In tropical forests, annual mortality can differ 20-60 fold among coexisting species \cite{Condit:2006}. This presumably reflects differences in the way trees from different species are constructed and function. Notably there exists a coordinated axis of life-history variation whereby individuals from species with higher potential growth rates also experience higher mortality \cite{Wright:2010fl}. At the same time the risk of death varies widely within species, with individuals growing slower or sitting lower in the light gradient at higher risk \cite{Kobe:1995tw, Wyckoff:2000un, Vieilledent:2010fv}. In other words mortality increases with growth rate across species, but decreases with growth rate within species. Hence the importance of separating hazards into those connected and unconnected with growth rate, and the need for integrating hazard estimation across and within species.

Although plants die from many causes, our approach described here partitions them into two fundamental categories: growth-dependent hazards and growth-independent hazards \cite{Keane:2001db, McDowell:2011dr}. Growth-independent hazards cause death irrespective of the tree's current growth rate, and may include windfall, fire, and certain types of pathogen or herbivore attack. These hazards are assumed to occur at some average rate (Fig 1a, left). Growth-dependent hazards cause slower-growing plants within each species to die at a faster rate, presumably because inferior carbon budgets make them unable to tolerate diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}. Many empirical studies suggest an exponential-like decline in risk with increased growth rate, $X_i(t)$ \cite{Kobe:1995tw, Wyckoff:2000un, Vieilledent:2010fv} (Fig 1a, centre). An individual's total risk of dying is then the sum of growth-independent plus growth-dependent hazards (Fig 1a, right). Biologically, the additive model implies that even fast-growing individuals experience some risk of death.

The additive hazards model has not previously been applied for empirical analysis, though it has been adopted in some vegetation models \cite{Moorcroft:2001ws}. In most previous empirical studies the underlying mathematical structure of statistical methods used implies responses similar to either left or centre panels of Fig 1a (see SI Appendix S1). In particular, the commonly used exponential models \cite{Kobe:1995tw, Wyckoff:2000un}, logistic regression \cite{Ruger:2011cv}, generalized linear models \cite{Visser_functional_2016}, and Cox proportional-hazard models \cite{Neumann_climate_2017} handle covariates in a manner that allows predicted risk to decline to zero for fast growing individuals (as in Fig 1a centre; SI Appendix S1). This makes these methods less likely to capture the full range of variation within and among species.

@@

Forest mortality rates are usually estimated from census data, where the size and alive vs dead status of individual trees is recorded at successive time points \cite{Anderson-Teixeira-2015a} (Fig 1b). From these data one can estimate $\lambda_i(t)$, a continuous, latent mortality rate or hazard rate for each individual, $i$, at time $t$. The probability an individual dies between times $t_1$ and $t_2$ results from the cumulative hazard they are exposed to between censuses\cite{Kleinbaum:2012tn}:
\begin{equation} \label{eq:mortality_model}
p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t\right),
\end{equation}
The observed survival outcome $S_{i, t_1\rightarrow t_2}$ (0 = alive, 1 = died) is then modelled as a binary draw. The challenge is then to estimate an appropriate form for the hazard rate, $\lambda_i(t)$.

Here we propose the following hazard function, incorporating individual-, species-, and census-period effects:
\begin{equation} \label{eq:model}
\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s}}^{\text{Growth independent}} + \overbrace{\alpha_{s} \, \textrm{e}^{-\beta_{s} \, X_{s,i}(t)}}^{\text{Growth dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
\end{equation}
Eq. \ref{eq:model} allows for additive growth-dependent and growth-independent effects, includes a negative exponential effect of the individual's growth rate (our measure of carbon budget), allows the overall mortality rate to vary among censuses via the random effect $\delta_{t}$, and also provides for species-level differences in the shape of $\lambda_i(t)$ via the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$. The species-level parameters are in turn modelled as a function of both species trait effects and species-level random effects, the latter of which accounts for residual species variation not attributed to traits.

\begin{figure}[!hb]
\centering
\includegraphics{figures/fig1.pdf}
\caption{Outline of methodology.
A) We consider three alternative hazard functions: 1) a growth-independent baseline hazard (A, left), 2) a growth-dependent hazard, where growth rate is a proxy for carbon budget (A, middle); and 3) a function that combines growth-dependent with growth-independent hazards (A, right). The parameters of the models are biologically interpretable: $\gamma$ is the baseline or growth-independent hazard; $\alpha$ defines the low growth rate (i.e. very marginal carbon budget) effect on the overall hazard; $\beta$ captures the sensitivity or curvature of the response to declining growth rate. In the most complex model (Panel A; right), $\alpha$ and $\beta$ capture growth-dependent hazards, while $\gamma$ captures growth-independent hazards that kill a plant, irrespective of its growth rate/carbon budget (e.g. windfall, fire). For each model form, we consider two alternative measures of growth, $X$ (basal area and stem diameter growth), and we allow for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
B) Our data consist of repeat measures of stem diameter ($D$) and status ($S$, 0 = alive or 1 = dead) for individual trees at specific census dates ($t_1$, $t_2$, $t_3$).
C) Each model's skill in predicting observed outcomes ($S$) is quantified via the log-loss function (eq. \ref{eq:logloss}). Lower logloss means higher predictive accuracy.
D) The predictive skill of alternative models was evaluated via 10-fold cross validation. The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$). Alternative models were fit 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing (1 fold; orange) and training (9 folds; green) data. Predictive capacity was assessed by averaging the log losses obtained from the 10 test data predictions.}
\label{fig1}
\end{figure}

To illustrate the proposed mortality model, we assess its ability to predict the survival status of \Sexpr{n_inds(BCI_training_full)} individual trees from \Sexpr{n_spp(BCI_training_full)} species, recorded over three 5-year census periods at Barro Colorado Island (BCI), Panama. Applying the additive model, we: i) estimate growth-dependent and growth-independent effects from long-term census data, ii) estimate species-level differences in these effects, and iii) assess how well species-level differences are predicted from three fundamental traits, previously shown to correlate with mortality: wood density, light demand, and maximum dbh \cite{king:2006va, Kraft:2010kq, Wright:2010fl, Ruger:2011cv}. These traits show only weak correlation with each other (r$^2$ \textless 0.13) and thus provide relatively independent measures of a species ecological strategy (SI Appendix, Fig. S6).

Using a likelihood-based penalty for incorrect prediction, we used 10-fold cross-validation to evaluate the skill of different models in their ability to predict outcomes in novel data (i.e. not used in model fitting; Fig. \ref{fig1}C-D; see Methods) \cite{Hooten:2015, Stone:1997tg}. In total \Sexpr{n_obs(BCI_training_full)} observations were used to fit these models. Evaluating models in this way is computationally expensive. Therefore rather than running all possible model formulations, we fitted models across six levels of increasing complexity, described below, and comparing their predictive skill. We then use the final model to assess the effects of species traits on each parameter and quantify their relative contributions on the overall model performance.

@@

\subsection*{Stage 1: Mortality over time}
We first assess whether average mortality rates varied among the three censuses, in other words whether a census effect is justified in Eq \ref{eq:model}. Comparing a model with a constant mortality rate $\gamma$ to a model where the constant is scaled by a census effect $\gamma\delta_{t}$, and examining the three 5-year census intervals, average mortality rates declined over time. The highest proportion of death occurred from 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_training_full, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_training_full, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_training_full, 3)}\%). Consequently, we observed a small increase in predictive skill when a census effect was included (leftmost two points under stage 1 in Fig. \ref{fig2}; see also Table.~\ref{table:1} for quantification of gain in predictive power under successive models).

\subsection*{Stage 2: Hazard functions}
Second, we assessed the predictive performance of three alternative hazard functions (Fig. \ref{fig1}A). The simplest form (Fig. \ref{fig1}A, left) assumes a constant hazard. The second form (Fig. \ref{fig1}A, middle) assumes the risk of dying declines towards an asymptote of zero as growth increases. The third "additive" model (Fig. \ref{fig1}BA right) is the sum of the previous two. In parallel, we investigated also which growth measure (increment of stem diameter or of stem area) predicts growth-dependent mortality better. Comparing the three hazard functions, the additive function significantly outperformed both the baseline (the best model in stage 1) and growth-dependent only hazard functions (right vs left of the four symbols under stage 2 in Fig. \ref{fig2}). Predictive skill was higher when using stem diameter growth compared to stem-area growth (Fig. \ref{fig2} stage 2, open vs closed symbols).

\subsection*{Stages 3-5: Species traits}
In stages 3-5, we take the additive model using dbh growth rate (the most predictive hazard function and growth rate measure from stage 2) and ask how addition of 1, 2 and 3 species-level traits improves model performance. Specifically, we introduce effects of the traits maximum dbh ($\psi$), light demand ($\upsilon$) and wood density ($\rho$) on the parameters $\alpha$, $\beta$ and $\gamma$ in Eq. \ref{eq:model}, at first one at a time (stage 3), then in 2-trait (stage 4) and 3-trait (stage 5) combinations. In general we found that including trait effects on these parameters decreased logarithmic loss, and thus, increased predictive accuracy (Fig.~\ref{fig2}). Allowing all three traits to have an effect on these parameters resulted in higher predictive capacity relative to models that only incorporated one or two of the traits.

\subsection*{Stages 6: Species random effects and trait effects}.

In order to examine how species differed in hazard functions and to understand the effects of species traits, we used the entire dataset to fit a final model containing all three species traits plus a species random effect to capture inter-specific variation not explained by traits. The increase in predictive accuracy afforded by traits (stage 5) was fairly small relative to including species random effects on each parameter, signifying that substantial inter-specific variation in mortality rates remains unexplained (Fig. \ref{fig2} Stage 6).

Each trait influenced tree mortality in different ways (Fig. \ref{fig3}-\ref{fig4}). Light demanding species were more susceptible to low growth rates (i.e. higher $\alpha$), had higher growth-independent hazards (i.e. higher $\gamma$) and exhibited shallower declines in growth-dependent hazards with increasing growth rates (i.e. lower $\beta$). Wood density had significant impacts on $\alpha$ and $\gamma$, such that species with high wood density were less susceptible to low growth rates and had lower growth-independent hazards. Species' maximum dbh only had significant effects on $\gamma$, whereby species capable of achieving larger dbh exhibited lower growth-independent hazards.

@@

\subsection*{Explanatory power}. 
The full model had an AUROC of \Sexpr{table1[[1,3]]}, indicating a fair ability to correctly discriminate observed deaths from survivals. Moreover, we found that the full model also explained \Sexpr{table1[[1,2]]*100}\% of the total variation in mortality observed among species.

Removal of the census random effect $\delta_t$ had little impact on either $r^2$ or AUROC. The removal of either wood density or light demand resulted in similar magnitude effects on both $r^2$ and AUROC. By contrast, the removal of maximum dbh had a slightly smaller effect on both metrics. Both $r^2$ and AUROC dropped substantially when species random effects were removed, further highlighting that maximum dbh, light demand and wood density explain only a small fraction of the overall inter-specific variation observed among species. When both species traits and species random effects were removed, the resulting a model was by definition unable to capture any variation observed among species, and thus had low discriminatory ability. Lastly, we found the removal of the growth-independent hazard term ($\gamma_s$) resulted in a lower $r^2$ relative to the removal of the growth-dependent hazard ($\alpha_s e^{-\beta_s X_{i}}$), indicating that the former hazard explained more inter-specific variation. However, when examining the removal of these hazards in terms of the model's ability to correctly distinguish survivors from those that died (AUROC), we found the opposite -- that the removal of the growth-dependent hazard resulted in substantially lower predictive ability.

@@

\section*{Discussion}
Our Bayesian framework coupled with cross validation revealed that the most predictive hazard model of tropical tree death was one which: 1) partitioned mortality into growth-dependent and growth-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) included the effects of species' maximum dbh, wood density and light demand; and 4) incorporated temporal variability. Further, rates of tropical tree mortality varied substantially between species and a species' maximum dbh, wood density and light demand explained only a fraction of the overall inter-specific variation.

There are several distinct advantages of modelling tree mortality as a function of two additive hazards: 1) The model explicitly accounts for the independent effects of the two fundamental types of hazards faced by plants; 2) it allows for the examination of the relative contributions each hazard provides to the overall hazard rate; 3) it allows for examination of how various factors affect each hazard term, and because of this; 4) allows ecological theories of how factors such as species traits affect plant mortality to be empirically tested using commonly collected data.

The findings of this study provide empirical support for dynamic vegetation models that predict tree mortality as a function of both growth-dependent and independent hazards \cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}. Regardless of growth measure (our proxy for carbon budget), incorporating both hazards significantly improved model predictive skill. This was because the growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stresses such as competition, parasites and herbivory. By contrast, the growth-independent hazard accounts for deaths caused by events that arise irrespective of an individual's carbon budget such as windthrow or lightning strike.

Partitioning mortality rates into growth-dependent versus growth-independent effects allowed us to estimate not only their relative contribution, but also how species traits affect each. Like many other studies \cite{King:1994vg, king:2006va, Ruger:2011cv, Kunstler:2016km}, our analyses highlighted the importance of light competition for tropical tree demographic rates. Specifically, removal of the growth-dependent hazard term resulted in a substantial reduction in the model's ability to correctly distinguish survivors from those that died (Table~\ref{table:1}). This suggests marginal carbon budgets are a major contributor to tree death on Barro Colorado Island.

Our analyses support previous research that advocated that wood density, species maximum size and light demand important explanatory variables of inter-specific variation in tropical rainforests \cite{king:2006va,Poorter:2006tq, Chave:2009iy}. Specifically, we show that light demanding species are not only more susceptible to both low growth rates and growth-independent hazards, but also exhibited shallower declines in growth-dependent hazards with increasing growth rates. By contrast, high maximum-dbh and high wood density species were found to be more tolerant of growth-independent hazards, corroborating evidence reported elsewhere \cite{Chave:2009iy, Kraft:2010kq}. Our analyses also support other work \cite{vanGelder:2006exa, Kunstler:2016km} showing that higher wood density species tend also to be more tolerant of low growth rates and better able to tolerate shade and competition.

Effects of species' wood density, maximum size and light demand are now being incorporated into mortality algorithms within many vegetation models in the effort to account for among-species variation \cite{Moorcroft:2001ws,Scheiter:2013ed}. However, the present analysis indicates that such traits are likely to capture only a small proportion of the overall inter-specific variation (Fig. \ref{fig4}, Table~\ref{table:1}). This means that dynamic vegetation models incoporating these traits are likely to underestimate the true variability in tropical rainforest mortality rates, and as a consequence underestimate carbon, water and nutrient dynamics of ecosystems \cite{Johnson-2016}.
